Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-11-12 Dail_Mad_ext.tex
model development sectino

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that resolves each of the problems outlined above. 
%simultaneous resolves each of the aforementioned problems with
%traditional state-space models, and is designed for simple count data.
%In the following section, we describe the DM model in its original
%form and explain how it resolves each of the deficiencies with
%standard state-space models. 
%In Section~\ref{sec:ext}, 
Our aim is to extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time series, such as excess zeros and unexplained random variation. 
%Specifically, we describe methods for accommodating
%excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
we evaluate the performance of the model using a simulation study by
analyzing data from the North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}.
%code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
%In Section~\ref{sec:app}, we evaluate the performance of the model
%extensions using a simulation study and by analyzing data from the
%North American Breeding Bird Survey (BBS), one of
%the most spatially and temporally extensive sets of count data on
%vertebrate populations \citep{robbins_etal:1986}. %The overarching aim
%of the paper is to provide ecologists with flexible and accessible means of
%addressing important questions related to the variation of abundance in
%space and time. 
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.
\begin{comment}
  The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error
using only spatially and temporally replicated count data. Their model
is an extension of the closed-population $N$-mixture model
\citep{royle:2004biom}, which was designed specifically to
address the problem of modeling spatial variation in abundance in the
  presence of observation error, specifically, when detection
  probability is less than one. The DM extension relaxes the
assumption of population closure and includes explicit
parameters describing population change as a first-order Markovian
  process. [more details]

[applications / potential etc... ] The DM model is relevant to a
vast number of problems confronting basic and applied ecologicalresearch. Answering questions about climate change etc...
require
spatially and temporally extensive datasets such as result from
the
  North American Breeding Bird Survey (BBS)
\citep{robbins_etal:1986}. However, most of the datasets of theappropriate scale are collected by volunteers and using
protocols
  that are not amenable to traditional approaches of modeling
  abundance and detection probability based on standard
capture-recapture methods. Instead they have, at best, simple
count
data for which, when it can be interpreted as counts of unique  individuals, the DM model can be applied.. Yadi ya

Although the DM model was designed explicitly for these
purposes,
the original formulation of the model makes strict
distributional
  assumptions and does not acknowledge many of the sources of
variation inherent to existing ecological time series data.  We  propose variations on these models that are likely to be more
realistic and useful in many cases. We demonstrate the
usefulness
  of these new models with simulated and real BBS data.

In this paper, we demonstrate how the DM model resolves each ofthese four limitations of conventional state-space models, and
we
extend the DM model in several important ways. First, we
demonstrate
  how classical population models can be embedded within the DM
model. Then we present methods for handling excess zeros as well
as
nuisance parameters such as random observer effects on detection probability. A simulation study
\end{comment}

@@

The DM model is an extension of the $N$-mixture model
\citep{royle:2004biom}, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate parameters associated with both abundance 
and the detection process, the original $N$-mixture model uses replicate
observations at each site. The observations are collected during sufficiently
short time intervals such that the population can safely be
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

\subsection{The Data}

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. 
%A site is ideally a
%well-defined region of a study area, such as a wetland or a
%patch of early-successional habitat, although it could be an arbitrarily
%defined region such as a randomly located survey plots. The time
%frame of the study is arbitrary, but in general
%it will be sufficiently long to allow for temporal variation in
%abundance. For instance, a primary sampling period could be a
%one month long breeding season, which could be surveyed once per
%year for $T$ years. If the a site is surveyed on $J$ occasions during a
%primary period, these are called secondary sampling occasions.
%In the case where no secondary sampling was conducted, i.e. $J=1$, 
Let $X_{i,t}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. 
%If secondary sampling periods were
%used, an extra dimension is added so that we have $X_{ijt}:
%j=1,\hdots,J$. 
Typically, we know $X_{i,t}$ will be lower than the
actual quantity of interest, abundance,
denoted $N_{i,t}$. In cases where detection probability is perfect,
abundance is observed directly such that $X_{i,t} \equiv N_{i,t}$.

@@

Like traditional state-space models, the DM model includes the
three conditionally related processes shown in Eq.~\ref{eq:ss1},
which correspond to (1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process %simply 
describes the the relationship between
abundance and the observed count data.

@@

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i,1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
%In the Poisson case, the equivalent expression of
%$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
%\begin{equation}
%[N_{i1} | \Lambda] =
%\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
%  \label{eq:N1br}
%\end{equation}
% Maybe we don't need the Poisson distriubtion equation? - Jeff

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process. %, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{i,t-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma(N_{i,t-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. Covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
%i.e. the model for abundance dynamics. This state
%model has allowed us to replace the initial abundance
%distribution and transition models of conventional state-space models, with
%alternatives that (1) allow for modeling spatial variation in
%abundance, (2) respect the discrete nature of the counts, and
%(3) have parameters that are estimable.
The one
remaining task is to replace the observation model
$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  X_{i,t} \sim \mathrm{Bin}(N_{i,t}, p)
  \label{eq:p1}
\end{equation}
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
Under this model, $\bm{p}$ is interpreted as the detection
probability parameter.

>>>>

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that resolves each of the problems outlined above. 
%simultaneous resolves each of the aforementioned problems with
%traditional state-space models, and is designed for simple count data.
%In the following section, we describe the DM model in its original
%form and explain how it resolves each of the deficiencies with
%standard state-space models. 
%In Section~\ref{sec:ext}, 
Our aim is to extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time series, such as excess zeros and unexplained random variation. 
%Specifically, we describe methods for accommodating
%excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
we evaluate the performance of the model using a simulation study by
analyzing data from the North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}.
%code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
%In Section~\ref{sec:app}, we evaluate the performance of the model
%extensions using a simulation study and by analyzing data from the
%North American Breeding Bird Survey (BBS), one of
%the most spatially and temporally extensive sets of count data on
%vertebrate populations \citep{robbins_etal:1986}. %The overarching aim
%of the paper is to provide ecologists with flexible and accessible means of
%addressing important questions related to the variation of abundance in
%space and time. 
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.

@@

The DM model is an extension of the 
\citet{royle:2004biom} $N$-mixture model, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate parameters associated with both abundance 
and the detection process, the original $N$-mixture model uses replicate
observations at each site. The observations are collected during sufficiently
short time intervals such that the population can safely be
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

%\subsection{The Data}

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. 
%A site is ideally a
%well-defined region of a study area, such as a wetland or a
%patch of early-successional habitat, although it could be an arbitrarily
%defined region such as a randomly located survey plots. The time
%frame of the study is arbitrary, but in general
%it will be sufficiently long to allow for temporal variation in
%abundance. For instance, a primary sampling period could be a
%one month long breeding season, which could be surveyed once per
%year for $T$ years. If the a site is surveyed on $J$ occasions during a
%primary period, these are called secondary sampling occasions.
%In the case where no secondary sampling was conducted, i.e. $J=1$, 
Let $X_{i,t}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. 
%If secondary sampling periods were
%used, an extra dimension is added so that we have $X_{ijt}:
%j=1,\hdots,J$. 
In general, $X_{i,t}$ will be lower than abundance, 
%actual quantity of interest, abundance,
%denoted 
$N_{i,t}$, but in cases where detection is perfect,
abundance is observed directly such that $X_{i,t} \equiv N_{i,t}$.

@@

Like traditional state-space models, the DM model includes 
three conditionally related processes corresponding to: 
% shown in Eq.~\ref{eq:ss1}, which correspond to 
(1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process %simply 
describes the the relationship between
abundance and the observed count data.

@@

%Recall that 
Conventional state-space models assume that the
distribution for the initial time period %,
%$[N_1 | \bm{\theta}]$, 
is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i,1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
%In the Poisson case, the equivalent expression of
%$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
%\begin{equation}
%[N_{i1} | \Lambda] =
%\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
%  \label{eq:N1br}
%\end{equation}
% Maybe we don't need the Poisson distriubtion equation? - Jeff

@@

The DM model assumes that abundance in time $t$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. %, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{i,t-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma(N_{i,t-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
%i.e. the model for abundance dynamics. This state
%model has allowed us to replace the initial abundance
%distribution and transition models of conventional state-space models, with
%alternatives that (1) allow for modeling spatial variation in
%abundance, (2) respect the discrete nature of the counts, and
%(3) have parameters that are estimable.
%The one
%remaining task is to replace the observation model
%$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
%more easily interpreted. 
The observation model adopted by \citet{dail_madsen:2011} is the same
binomial model proposed by \citet{royle:2004biom}: % assumes that individuals are missed due to imperfect
%detection. The simplest model for imperfect detection is
\begin{equation}
  X_{i,t} \sim \mathrm{Bin}(N_{i,t}, p)
  \label{eq:p1}
\end{equation}
where $p$ is the probability of detecting each individual. Variation
in detection probability can be modeled as a function of site-specific
or occasion-specific covariates using, for example, a logit-linear model.
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
%Under this model, $\bm{p}$ is interpreted as the detection
%probability parameter.
